Leader functions and qualities across 59 cultures: an exploratory analysis

### 2019-12-03 leadership_across_cultures_contexts.Rmd
Added more ethnographic examples in discussion section.

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. These variables were *not* determined prior to the study but were instead identified and coded during several passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence stype, and sex of the leader.

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the surpacommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Adamans in LOCATION, by AUTHOR [@az02-001] (45). (reproducing AUTHOR), who discussed the roles of a "head chief" from a permnant residential location and their marginal influence over other, local residential leaders:

@@

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013], exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratigied agro-pastoralists group, DESCIRBE, AUTHOR [@mo04-021] (p. 58) notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain imporant leadership positions within the groups na drepresent the group to govermental forces. AUTHOR [@aa01-062] (p. 173) describes that in these local 'clubs', "the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. AUTHOR [@oc06-034] (p. 236) in describing kin group leaders among the Iban in LOCATION provides example of this suite of traits in leaders:

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, AUTHOR [@nf06-061] (p. 411) descirbeds that, "If there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful categories. Variables were clustered together if they were positively correlated. Leader qualities formed two top-level clusters, which we identify as a *Material, social, and reproductive success* cluster, and a *Social orientation* cluster. See Figure \@ref(fig:cluster)A. Several of traits often cooccur. Among the Trobriands horticulturalists in LOCATION, AUTHOR [@ol06-037] (p. 47 ) describes relationships between plogynous marriages, increased wealthm, and widespread social contacts:

@@

Most of the leader quality variables were found in the *Social orientation* cluster, which had two smaller and more weakly supported subclusters that we identified as *Ability to provide social benefits* and *Ability to impose social costs*. The social benefits subcluster contained the *Knowledgeable/intelligent* and *Experienced/accomplished* variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and these variables are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport), a point to which we shall return. As brief examples of the very many in the ethngorraphic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them." [fo07-056] (p. 85); and among the Highland Scots, AUTHOR [es10-003] (p. 147) notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit heirachy in the context of cultural transitions. AUTHOR [@nm09-056] (p. 183) describes Minnie Kellog (born in 1880), a land-claims activist, writer, and perhaps the most famous female leader among the Iriquous:

@@

The leader function variables formed two major clusters, one that we identified as "Prosociality" and one that we identified as "Management". The prosocial cluster included *Distribute resources*, *Protection*, and *Prosocial investment*, which are functions that are commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absenece of institutionalized leadership roles, which charachterize many of the relatively small-scale populations in this sample, leaders often maintain social influence via continued community support and the demands of followers can shape the functions of leaders to meet local needs.

The management subcluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991]. Other variables in this subcluster indicated more generic group strategizing, such as *Strategic planning* and *Group representative* [@de_cremer_intergroup_2002]. In the absenece of state-instiutions local leaders are often tasked with managing various economic systems and providing social services.

The larger organization subcluster included variables such as *Organize cooperation*, *Ritual functions*, and *Military command*. This subcluster aligns with evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [@hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019].

@@

Intergenerational transmission; The fact that variables indicating success in these three domains were correlated supports the...

The strong support for material and social benefits assocaited with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequlity [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or instituionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." Theses results suggests anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporates and pay close attention to group processes and social strategies associated with leadership and followership.

>>>>

This exploratory study extracted the qualities, functions, benefits, and costs of leadership from the ethnographic record. These variables were *not* determined prior to the study but were instead identified and coded during several passes through `r nrow(text_records)` texts on leadership obtained from the eHRAF (see Methods section). Our results therefore represent the broadest and most comprehensive view of the qualities of leaders (i.e., the types of people that occupy leadership roles), the functions of leaders (i.e., what leaders do), and the costs and benefits of leadership for both leaders and followers across a diverse and culturally representative sample of the ethnographic record. These results reveal universal dimensions of human leadership while also suggesting important variation by group type, subsistence type, and sex of the leader.

@@

Leadership in residential subgroups (e.g., hunting party, work group) was relatively common among hunter-gatherers but relatively rare among agriculturists. Leadership at the surpacommunity level, in contrast, was relatively rare among hunter-gatherers but quite common among agriculturists. The only example of supracommunity leadership among hunter-gatherers was from the Andamanese, a group of populations of Negrito hunter-gatherers of the Andaman Islands  by @az02-001: p. 45, who discussed the roles of a "head chief" from a permanent residential location and their marginal influence over other, local residential leaders:

@@

Hunter-gatherers and pastoralists, populations more likely to experience high residential mobility [@wild_making_2019; @rigby_persistent_1985; @kelly_foraging_2013], exhibited greater relative frequencies of leadership within residential groups and kin groups. Pastoralists, however, rely in part on a highly defensible resource base (i.e., livestock) and exhibit leadership beyond the community more so than hunter-gatherers and on par with horticulturalists. Pastoral and horticultural populations often experience elevated rates of inter-group conflict relative to mobile hunter-gatherers and these results are consistent with a relationship between increased inter-group conflict and more elaborated political leadership systems [@ember_warfare_1992]. For example, among the Somali, a politically complex and socially stratified agro-pastoralists group distributed across the Horn of Africa, @mo04-021 (p. 58) notes that, "the Sultan does not lead the army of the tribe in war. He names a chief of the armed forces."

Figure \@ref(fig:grouptype) is consistent with a negative relationships between political leadership at the community leadership or above and leadership in kin groups or residential groups. One possible interpretation is that as political structures become more institutionalized and economic systems transcend the household, leadership within households and residential groups diminishes It is interesting, however, that even among the agricultural populations that are characterized by leadership systems beyond the community level, leadership within residential and family group persists. For example, among rural Korean populations a chairwomen of local organizations maintain important leadership positions within the groups and represent the group to governmental forces. @aa01-062 (p. 173) describes that in these local 'clubs', "the procedure of discussion and decision-making is not based on an idea of democracy but rather on that of unanimity. In this context, the ‘leader’ does not have to be a person of marked influence in the decision-making process. She is not a figurehead but rather a middleman in the meeting, functioning, in effect, as a go-between for the group. There is no despotic idea in the context of a total consensus in this essentially egalitarian society. The chairwoman stands as a middleman between the government and the village. Although the government may call her a ‘leader’ in the village, she properly does not occupy this role."

@@

Across a broad range of cultures and social contexts leaders are nearly universally high status males who are described as knowledgeable or intelligent and who are experienced and have a demonstrated history of previous accomplishments or special skills. They most frequently resolve conflicts within their respective groups while also providing important social functions, organizing cooperation among followers, providing counsel or input in followers' affairs, and distributing resources. @oc06-034 (p. 236) in describing kin group leaders among the Iban, small-scale highland horticulturalists primarily in Borneo and Sarawak, provides example of this suite of traits in leaders:

@@

Several functions and qualities that are theoretically important were relatively rare in the ethnographic record. In the literature on leadership among non-human animals, for example, leaders commonly initiate group movement [@couzin_effective_2005], and "movement" was one of four leadership dimensions that @smith_leadership_2016 identified as comparable among humans and non-human animals, yet movement or migration was identified as a leader function in only `r movement_pct`% of cultures, of which the vast majority (71%) were hunter-gatherers. Among the Blackfoot in the North American plains for example, @nf06-061 (p. 411) describes that, "If there was a dispute as to whether the camp should move north or south, the chief might present his arguments, gain support, and then say that he was going south. He did not order the others to follow, but he knew that they probably would go with him."

@@

We performed hierarchical cluster analysis on our leader quality and leader function variables to determine if these variables formed theoretically meaningful categories. Variables were clustered together if they were positively correlated. Leader qualities formed two top-level clusters, which we identify as a *Material, social, and reproductive success* cluster, and a *Social orientation* cluster. See Figure \@ref(fig:cluster)A. Several of traits often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 (p. 47 ) describes relationships between polygnyous marriages, increased wealth, and widespread social contacts:

@@

Most of the leader quality variables were found in the *Social orientation* cluster, which had two smaller and more weakly supported subclusters that we identified as *Ability to provide social benefits* and *Ability to impose social costs*. The social benefits subcluster contained the *Knowledgeable/intelligent* and *Experienced/accomplished* variables that have been emphasized by the *Prestige* theory [@henrich_evolution_2001; @cheng_two_2013] and the more general *Computational services* theory [@hagen_leadership_2019; @garfield_evolutionary_2019], and these variables are among the most strongly supported in the ethnographic record (Figure \@ref(fig:varsupport), a point to which we shall return. As brief examples of the very many in the ethnographic record, among the Azande, "though the most knowledgeable witch-doctors tend to become leaders, there is no institutionalized organization or leadership among them." [fo07-056] (p. 85); and among the Highland Scots, es10-003 (p. 147) notes, "on the whole, their long ties with place and people, coupled with education, made them useful leaders of a community in which leaders were needed to adapt life to a changing economy." Such exceptionally talented leaders, however, also face unique challenges, particularly within more egalitarian societies which emphasizes personal autonomy and resit hierarchy in the context of cultural transitions. @nm09-056 (p. 183) describes Laura Cornelius Kellogg (also known as Minnie and Wynnogene, 1880-1947), a land-claims activist, writer, and perhaps the most famous female leader among the Iroquois:

@@

The leader function variables formed two major clusters, one that we identified as "Prosociality" and one that we identified as "Management". The prosocial cluster included *Distribute resources*, *Protection*, and *Prosocial investment*, which are functions that are commonly highlighted in the anthropological literature on leadership in nonindustrial societies [@sahlins_poor_1963; @johnson_organizational_1982; @service_origins_1975; @boehm_egalitarian_1993; @boehm_hierarchy_1999; @johnson_evolution_1987]. In the absence of institutionalized leadership roles, which characterize many of the relatively small-scale populations in this sample, leaders often maintain social influence via continued community support and the demands of followers can shape the functions of leaders to meet local needs.

The management cluster included many more variables, which were grouped into two subclusters that we identified as "Mediation" and "Organization". The mediation subcluster included variables such as *Resolve conflict* (the most widely supported function in our sample) and *Provide counsel* (also widely supported), which again are functions that are commonly noted in the anthropological literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991], and concerning conflict resolution specifically, widespread among nonhuman animals as well [@smith_leadership_2016; @smith_obstacles_2018; @wilson_sociobiology:_1980]. Among the Mbuti, hunter-gatherers in the Congo Basin known for strong cultural values of personal autonomy and egalitarianism, @fo04-001 (p. 185) describes the nature of conflict resolution processes in the absence of strong authority:

>The elder settles numerous disputes, both within and without his hunting band. Such disputes mainly concern marital disagreements, Schebesta says. The elders often settle controversies by delivering speeches during the evening, leading frequently to noisy debates between the elders of different bands or lineages. Outside the lineage context, the elder is not in any way different from the others; there are none of the trappings of villager chieftainship. He delivers his decisions without hesitancy and is obeyed in the same way within his lineage. His power is limited, however, by custom and tradition as well as by context. He may neither ignore nor alter traditional practices.

Other variables in this subcluster indicated more generic group strategizing, such as *Strategic planning* and *Group representative* [@de_cremer_intergroup_2002]. In the absence of state-institutions local leaders are often tasked with managing various economic systems and providing social services.

The larger organization subcluster included variables such as *Organize cooperation*, *Ritual functions*, and *Military command*. This subcluster aligns with the many evolutionary theories of leadership that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; for review, see @garfield_evolutionary_2019]. In the ethnographic record such functions includes controlling and facilitating economic productivity and ensuring collective actions do not overlly conflict with individual or household-level goals, such as among the Tukano of the Braziallian Amazon, where @sq19-018 (p. 127) notes:

>Wanano chieftainship entails control over three major aspects of village life: fishing resources, labor, and ritual. A mahsa wami must hold sway over all three areas to lead effectively. Inability to harvest and redistribute fish, for example, will make it difficult for the mahsa wami to mobilize and coordinate labor, Control, then, implies the obligation to redistribute. The chief mobilizes and coordinates communal labor, carefully balancing collective projects against individual work such as gardening, so that collective and individual labor do not conflict. I observed several kinds of community projects in Wanano villages, including the construction and maintenance of houses, the clearing and burning of gardens, and the intensive preparations for intersib po?oa exchange ceremonies.

@@

The strong support for material and social benefits associated with leadership (Figure \@ref(fig:varsupportcb)) and the colinearity of measures associated with material wealth, social capital, and reproductive success (Figure \@ref(fig:cluster)A), suggests processes of leadership and followership might be strong drivers of the emergence and intergenerational transmission of inequality [@bowles_emergence_2010; @mattison_evolution_2016]. Various sources of material wealth and strategic or institutionalized processes of nepotism are widespread cross-culturally, including among many societies often described as "egalitarian." Theses results suggests anthropologists and economists investigating the emergence and evolution of inequality, including material, social, and reproductive dimensions, should incorporates and pay close attention to group processes and social strategies associated with leadership and followership.
